---
title: "Appendix S2"
author: ""
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Bringing circuit theory into spatial occupancy models to assess landscape connectivity**, Maëlis Kervellec, Thibaut Couturier, Sarah Bauduin, Delphine Chenesseau, Pierre Defos du Rau, Nolwenn Drouet-Hoguet, Christophe Duchamp, Julien Steinmetz, Jean-Michel Vandel, Olivier Gimenez
  
   
We present in the following document the code to implement dynamic occupancy models using NIMBLE (de Valpine et al., 2017). Reproducible examples with simulated data are available on Zenodo XX in the `SpatialOccupancyModels` folder. 

```{r}
require(nimble)
```

# 1. Dynamic occupancy model  
You can find here the nimble code to run standard dynamic occupancy models. For more information on the model see Mackenzie et al. (2017). 
  
```{r, eval = FALSE}
mackenzie.model <- nimbleCode({
  
  # Priors: ecological parameters
  psi1 ~ dunif(0, 1)  # initial occupancy probability
  gamma ~ dunif(0, 1) # colonization probability
  phi ~ dunif(0, 1)   # persistence probability
  p ~ dunif(0, 1)     # detection probability
  
  # ecological submodel
  for (i in 1:nsites){
    z[i,1] ~ dbern(psi1)
    
    for (t in 2:nseasons){
      muZ[i,t]<- z[i,t-1] * phi + (1 - z[i,t-1]) * gamma
      z[i,t] ~ dbern(muZ[i,t])
    }
  }
  
  # observation model
  for (i in 1:nsites){
    for (t in 1:nseasons){
      y[i,t] ~ dbinom(size = nsurveys, prob = z[i,t] * p)
    }
  }
})
```


# 2. Spatial dynamic occupancy model  
You can find here the nimble code to run spatial dynamic occupancy models using euclidean distance. For more information on the model see Chandler et al. (2015). 
  
```{r, eval = FALSE}
## Define distance -------------------------------------------------------------
d <- AHMbook::e2dist(xy,xy) # euclidean distance between all pairs of sites
distSq <- (d^2)/sd(d^2)

## Model -----------------------------------------------------------------------
euclidean.model <- nimble::nimbleCode({
  
  # Priors: ecological parameter
  psi1 ~ dunif(0, 1)   # first session occupancy probability
  gamma0 ~ dunif(0, 1) # baseline colonization probability
  sigma ~ dgamma(shape = 1, rate = 1) # scale parameter
  phi ~ dunif(0, 1)   # persistence probability
  p ~ dunif(0, 1)     # detection probability
  
  # ecological submodel
  for (i in 1:nsites){
    z[i,1] ~ dbern(psi1)
    
    for(t in 2:nseasons) {
      for(n in 1:nsites) {
        # pairwise colonization probability
        gammaDistPairs[i,n,t-1] <- 1 - gamma0 * exp(-distSq[i,n] / (2 * sigma^2)) * z[n,t-1]
      }
      # colonization probability
      gamma[i,t-1] <- 1 - prod(gammaDistPairs[i,1:nsites,t-1])
      ## Pr(z=1|z=1) = 1 - Pr(extinction)*Pr(not colonized) = 1 - ((1-phi)*(1-gamma))
      muz[i,t-1] <- gamma[i,t-1] * (1 - z[i,t-1]) + phi * z[i,t-1]
      z[i,t] ~ dbern(muz[i,t-1])
    }
  }
  # observation model
  for (i in 1:nsites){
    for (t in 1:nseasons){
      y[i,t] ~ dbinom(size = nsurveys, prob = z[i,t] * p)
    }
  }
})
```

# 3. Spatial dynamic occupancy model accomodating Least Cost Path distance  
You can find here the nimble code to run spatial dynamic occupancy models using least cost path distance. For more information on the model see Howell et al. (2018). 
  
```{r, eval = FALSE}
## Define distance -------------------------------------------------------------
leastcostpath <- function(alpha){
  # Compute least cost path between a set of points given a resistance parameter 
  # alpha and a resistance surface rcov
  
  # alpha is a scalar
  # rcov is a RasterLayer
  # xy is a matrix of dims nsites x 2
  
  # obtain resistance surface
  cost <- exp(alpha * rcov)
  # compute conductances among neighbours 
  # probability of transition from one cell to another
  tr <- gdistance::transition(x = cost, 
                              transitionFunction = function(x) 1/mean(x),
                              directions = 8) # class TransitionLayer
  # adjust diag.conductances
  trCorrC <- gdistance::geoCorrection(x = tr, 
                                      type = "c", 
                                      multpl = FALSE, 
                                      scl = FALSE) # class TransitionLayer
  # compute ecological distance
  d <- gdistance::costDistance(x = trCorrC,
                               fromCoords = xy, 
                               toCoords = xy) # class Matrix
  d^2
}

# Create the nimble function to compute distance between sites 
LCP <- nimbleRcall(function(alpha = double(0)){},
                   Rfun = 'leastcostpath',
                   returnType = double(2))

## Model -----------------------------------------------------------------------
lcp.model <- nimble::nimbleCode({
  
  # Priors: ecological parameter
  psi1 ~ dunif(0, 1)   # first session occupancy probability
  gamma0 ~ dunif(0, 1) # baseline colonization probability
  alpha2 ~ dunif(-5, 5) # resistance parameter
  sigma ~ dgamma(shape = 1, rate = 1) # scale parameter
  phi ~ dunif(0, 1)   # persistence probability
  p ~ dunif(0, 1)     # detection probability
  
  # computation of distances for a given resistance
  distSq[1:nsites,1:nsites] <- LCP(alpha2) 
  distSqreduced[1:nsites,1:nsites] <- distSq[1:nsites,1:nsites] / sd(distSq[1:nsites,1:nsites]) 
  # reduced to help convergence
  
  # ecological submodel
  for (i in 1:nsites){
    z[i,1] ~ dbern(psi1)
    
    for(t in 2:nseasons) {
      for(n in 1:nsites) {
        # pairwise colonization probability
        gammaDistPairs[i,n,t-1] <- 1 - gamma0 * exp(-distSqreduced[i,n] / (2 * sigma^2)) * z[n,t-1]
      }
      # colonization probability
      gamma[i,t-1] <- 1 - prod(gammaDistPairs[i,1:nsites,t-1])
      muz[i,t-1] <- gamma[i,t-1] * (1 - z[i,t-1]) + phi * z[i,t-1]
      z[i,t] ~ dbern(muz[i,t-1])
    }
  }
  # observation model
  for (i in 1:nsites){
    for (t in 1:nseasons){
      y[i,t] ~ dbinom(size = nsurveys, prob = z[i,t] * p)
    }
  }
})
```

# 3. Spatial dynamic occupancy model accomodated with commute distance  
You can find here the nimble code to run spatial dynamic occupancy models using commute distance from circuit theory. More information on the model are presented in the paper.  
  
```{r, eval = FALSE}
## Define distance -------------------------------------------------------------
circuitdistance <- function(alpha){
  # alpha is a scalar
  # rcov is a RasterLayer
  # xy is a matrix of dims nsites times 2
  # obtain resistance surface
  cost <- exp(alpha * rcov)    
  # compute conductances among neighbours 
  # probability of transition from one cell to another
  tr <- gdistance::transition(x = cost, 
                              transitionFunction = function(x) 1/mean(x),
                              directions = 8) # class TransitionLayer
  # adjust diag.conductances
  trCorrC <- gdistance::geoCorrection(x = tr, 
                                      type = "r", # for randomwalk
                                      scl = TRUE) # class TransitionLayer 
  
  #Circuit theory = random walk with theta = 0
  circuitDist <- gdistance::commuteDistance(x = trCorrC, coords = xy) %>%  
    as.matrix() 
  dist <- circuitDist
  return(dist^2)
}

CircuitDistance <- nimbleRcall(function(alpha = double(0)){},
                               Rfun = 'circuitdistance',
                               returnType = double(2))

## Model -----------------------------------------------------------------------
circuit.model <- nimble::nimbleCode({
  
  # Priors: ecological parameter
  psi1 ~ dunif(0, 1)   # first session occupancy probability
  gamma0 ~ dunif(0, 1) # baseline colonization probability
  alpha2 ~ dunif(-5, 5) # resistance parameter
  sigma ~ dgamma(shape = 1, rate = 1) # scale parameter
  phi ~ dunif(0, 1)   # persistence probability
  p ~ dunif(0, 1)     # detection probability
  
  # computation of distances for a given resistance
  distSq[1:nsites,1:nsites] <- CircuitDistance(alpha2)
  distSqreduced[1:nsites,1:nsites] <- distSq[1:nsites,1:nsites] / sd(distSq[1:nsites,1:nsites])
  
  # ecological submodel
  for (i in 1:nsites){
    z[i,1] ~ dbern(psi1)
    
    for(t in 2:nseasons) {
      for(n in 1:nsites) {
        # pairwise colonization probability
        gammaDistPairs[i,n,t-1] <- 1 - gamma0 * exp(-distSqreduced[i,n] / (2 * sigma^2)) * z[n,t-1]
      }
      # colonization probability
      gamma[i,t-1] <- 1 - prod(gammaDistPairs[i,1:nsites,t-1])
      muz[i,t-1] <- gamma[i,t-1] * (1 - z[i,t-1]) + phi * z[i,t-1]
      z[i,t] ~ dbern(muz[i,t-1])
    }
  }
  # observation model
  for (i in 1:nsites){
    for (t in 1:nseasons){
      y[i,t] ~ dbinom(size = nsurveys, prob = z[i,t] * p)
    }
  }
})
```

# References

Chandler, R. B., E. Muths, B. H. Sigafus, C. R. Schwalbe, C. J. Jarchow, and B. R. Hossack. 2015. Spatial occupancy models for predicting metapopulation dynamics and viability following reintroduction. Journal of Applied Ecology 52:1325–1333.

Howell, P. E., E. Muths, B. R. Hossack, B. H. Sigafus, and R. B. Chandler. 2018. Increasing connectivity between metapopulation ecology and landscape ecology. Ecology 99:1119–1128.

MacKenzie, D. I., J. D. Nichols, J. A. Royle, K. H. Pollock, L. Bailey, and J. E. Hines. 2017. Occupancy Estimation and Modeling: Inferring Patterns and Dynamics of Species Occurrence. Second edition. Elsevier.

de Valpine, P., D. Turek, C. J. Paciorek, C. Anderson-Bergman, D. T. Lang, and R. Bodik. 2017. Programming with models: writing statistical algorithms for general model structures with NIMBLE. Journal of Computational and Graphical Statistics 26:403–413.

