---
title: "Spatial occupancy models from simulated occupancy data"
author: "Maëlis Kervellec"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(nimble)
library(patchwork)

theme_set(theme_light(base_size = 14))
```

In this script, we present four occupancy models fitted to simulated data. Among the four models, three incorporate spatial components by utilizing different distance metrics: Euclidean distance, least-cost path (LCP) distance, and commute distance. 


# 1. Simulate data 
## 1.1 Define the resistance surface and occupancy surface 

First, we choose a structured landscape from which we are going to define the resistance surface and sampling sites. To define barriers we define a study area around a small portion of the river network in Aveyron in France. The resistance surface is computed as the distance to the rivers, thereafter called barriers. We kept a 2 pixels buffer to define the 11 by 11 sampling grid. 

```{r}
source("SpatialOccupancyModels/functions.R")
load("SpatialOccupancyModels/barriers.RData")

# Define extend of the study area ----------------------------------------------
box <- tibble(X = c(553055, 553055,565055, 565055),
              Y = c(1909587, 1921587, 1909587, 1921587)) %>%
  sf::st_as_sf(coords = c("X","Y"), crs = 27572) %>%
  sf::st_make_grid() %>%
  sf::st_union()

# Make a grid ------------------------------------------------------------------
grid <- sf::st_make_grid(box, cellsize = 250, square = TRUE) %>%
  sf::st_sf() %>%
  dplyr::mutate(id = 1:nrow(.))

# Compute the distance to the barrier ------------------------------------------
nearest_barrier <- sf::st_nearest_feature(grid, barriers)
dist_barrier <-  sf::st_distance(grid, barriers$geometry[nearest_barrier], by_element = TRUE)
grid <- grid %>%
  dplyr::mutate(dist_barrier = dist_barrier)

# Rasterize --------------------------------------------------------------------
coords <- grid %>%
  sf::st_centroid() %>%
  sf::st_coordinates()

rcov <- data.frame(x = coords[,1],
                   y = coords[,2],
                   dist_barrier = as.numeric(dist_barrier))%>%
  raster::rasterFromXYZ(crs = 27572)

rcov@data@values <- rcov@data@values/sd(rcov@data@values) 

mapview::mapview(barriers) + mapview::mapview(rcov)

# Occupancy sites --------------------------------------------------------------
box_sites  <- tibble(X = c(553055+500, 553055+500, 565055-500, 565055-500),
                     Y = c(1909587+500, 1921587-500, 1909587+500, 1921587-500)) %>%
  sf::st_as_sf(coords = c("X","Y"), crs = 27572) %>%
  sf::st_make_grid() %>%
  sf::st_union()

grid_sites <- sf::st_make_grid(box_sites, cellsize = 1000, square = TRUE) %>%
  sf::st_sf() %>%
  dplyr::mutate(id = 1:nrow(.))
mapview::mapview(rcov)+ mapview::mapview(grid_sites)

xy <- sf::st_coordinates(sf::st_centroid(grid_sites))[,c("X","Y")] # occupancy sites coordinates
```
In total the resistance surface is composed of `r nrow(grid)` pixels, and the occupancy surface contains `r nrow(grid_sites)` sites. 

## 1.2 Compare least-cost path distance to commute distance

We compare LCP and commute distances. We choose a source pixel in the landscape and then compute the two distances to all pixel in the landscape, for a given resistance parameter. 
```{r}
alpha <- -0.5
leastcostpath_1 <- LCP(alpha)
circuitdistance_1 <- CircuitDistance(alpha)

# Correlation
i <- 60
cor(leastcostpath_1[i,],circuitdistance_1[i,])

# Map the 2 resistance raster from cell (1,1), with the 2 methods
A <- data.frame(x=xy[,1],y=xy[,2],value=leastcostpath_1[i,])
B <- data.frame(x=xy[,1],y=xy[,2],value=circuitdistance_1[i,])

Palette <- colorRampPalette(c("#e0e1dd","#778da9","#415a77", "#1b263b", "#0d1b2a"))

Source <- data.frame(x=xy[i,1],y=xy[i,2])
mapA <- ggplot()+
  geom_raster(data = A, mapping = aes(x=x, y=y, fill= value))+
  scale_fill_gradientn(colors = Palette(5)) +
  geom_point(data = Source, mapping = aes(x=x,y=y), col = "#e63946") +
  ggtitle("Least cost path distance, \nalpha = -0.5 ")

mapB <- ggplot()+
  geom_raster(data = B, mapping = aes(x=x, y=y, fill= value))+
  scale_fill_gradientn(colors = Palette(5)) +
  geom_point(data = Source, mapping = aes(x=x,y=y), col = "#e63946") +
  ggtitle("Commute distance, \nalpha = -0.5 ")

mapA + mapB
```

## 1.3 Simulate occupancy data 

We simulate detection history data in this landscape. We assume that the true model is the model with commute distance. 
```{r}
# Set variable -----------------------------------------------------------------
M <- nrow(xy) # number of sites
J <- 4 # number of sampling occasions
T <- 3 # number of seasons
psi1 <- 5/M # initial occupancy probability
alpha2 <- -2 # resistance parameter for Circuit
sigma <- 0.1 # scale parameter
gamma0 <- 0.2 # baseline colonization probability
phi <- 0.9 # persistence probability
p <- 0.5 # detection probability

# Simulate detection/non-detection data ----------------------------------------
Data <- Simulate_occ_data(M, J, T, psi1, alpha2, sigma, gamma0, phi, p, method = "Circuit") # if method = "LCP", you can simulate detection/non-detection data according to a spatial occupancy model with LCP distance

# Plot true presence/absence at each season ------------------------------------ 
grid_sites <- grid_sites %>%
  dplyr::mutate(nbDet =  rowSums(Data[["y"]][,,1]),
                Presence1 = Data[["z"]][,1],
                Presence2 = Data[["z"]][,2],
                Presence3 = Data[["z"]][,3])

grid_sites %>%
  ggplot2::ggplot() +
  ggplot2::geom_sf(aes(fill = Presence1))
grid_sites %>%
  ggplot2::ggplot() +
  ggplot2::geom_sf(aes(fill = Presence2))
grid_sites %>%
  ggplot2::ggplot() +
  ggplot2::geom_sf(aes(fill = Presence3))
```
# 2. Fit spatial occupancy models 

In this section, we present 4 occupancy models and fit them to our simulated data using NIMBLE. 

## 2.1 No distance 

The first model is the standard occupancy model defined by Mackenzie et al. (2017). This non-spatial model accounts for detection and dynamic colonisation/extinction of sites. 

```{r}
# Data and constants 
my.constants <- list(nsites = M, 
                     nsurveys = J, 
                     nseasons = T)
my.data <- list(y = apply(Data[["y"]], MARGIN = c(1,3), FUN = sum))

zst <- array(1, dim = c(M, T))
initial.values <- function(){ list(z = zst, psi1 = runif(1,0,1), gamma = runif(1,0,1), phi = runif(1,0,1), p = runif(1,0,1))}

# Initialization and compilation
occupancy <- nimbleModel(code = mackenzie.model, 
                         data = my.data,
                         constants = my.constants,
                         inits = initial.values(),
                         calculate = FALSE) # first tip

Coccupancy <- compileNimble(occupancy)

occupancyConf <- configureMCMC(occupancy,
                               enableWAIC = TRUE)

occupancyMCMC <- buildMCMC(occupancyConf, useConjugacy = FALSE) # second tip

CoccupancyMCMC <- compileNimble(occupancyMCMC, 
                                project = occupancy)

# Run Nimble
samples <- runMCMC(mcmc = CoccupancyMCMC, 
                   niter = 10000, 
                   nburnin = 5000,
                   nchains = 2,
                   WAIC = TRUE)

# Plots
mcmc.output.mackenzie <- samples
mcmc.output.mackenzie$WAIC
MCMCvis::MCMCtrace(object = mcmc.output.mackenzie$samples,
                   pdf = FALSE, # no export to PDF
                   ind = TRUE, 
                   params = c("gamma", "psi1", "phi", "p"),
                   gvals = c(NA, psi1, phi, p))
```


## 2.2 Euclidean distance  

This spatial occupancy model was defined by Chandler et al. (2015). This hierarchical model extend the previous one by integrating an Euclidean distance in the colonization process. 

```{r}
# Compute distance between sites 
d <- AHMbook::e2dist(xy,xy) # euclidean distance between all pairs of sites
distSq <- (d^2)/sd(d^2) # Need distance squared in half-normal kernel, and reduce to avoid large values and help convergence

# Data and constants
my.constants <- list(nsites = M, 
                     nsurveys = J, 
                     nseasons = T,
                     distSq = distSq)

my.data <- list(y = apply(Data[["y"]], MARGIN = c(1,3), FUN = sum))

zst <- array(1, dim = c(M, T))

initial.values <- function(){ list(z = zst, 
                                   psi1 = runif(1,0,1), 
                                   sigma = rgamma(1, shape = 1, rate = 1),
                                   gamma0 = runif(1,0,1),
                                   phi = runif(1,0,1), 
                                   p = runif(1,0,1))}

# Initialization and compilation
occupancy <- nimble::nimbleModel(code = euclidean.model, 
                                 data = my.data,
                                 constants = my.constants,
                                 inits = initial.values(),
                                 calculate = FALSE) # first tip

Coccupancy <- nimble::compileNimble(occupancy)

occupancyConf <- nimble::configureMCMC(occupancy,
                                       enableWAIC = TRUE)

occupancyMCMC <- nimble::buildMCMC(occupancyConf, useConjugacy = FALSE) # second tip

CoccupancyMCMC <- nimble::compileNimble(occupancyMCMC, 
                                        project = occupancy)

# Run Nimble
samples <- nimble::runMCMC(mcmc = CoccupancyMCMC, 
                           niter = 10000, 
                           nburnin = 5000,
                           nchains = 2,
                           WAIC = TRUE)

# Plots
mcmc.output.euclidean <- samples
mcmc.output.euclidean$WAIC
MCMCvis::MCMCtrace(object = mcmc.output.euclidean$samples,
                   pdf = FALSE, # no export to PDF
                   ind = TRUE, 
                   params = c("gamma0", "sigma", "psi1", "phi", "p"),
                   gvals = c(gamma0, sigma, psi1, phi, p))
```

## 2.3 Least-cost path distance 

This spatial occupancy model was defined by Howell et al. (2018). This model account for resistance of the landscape in between sites movements. To do so, it include a LCP distance in the colonization process, which allows for an explicit estimation of connectivity. 

### 2.3.1 Compute cached matrix 

To avoid to compute LCP distance at each MCMC iteration, we precalculate distance between sites for a set of alpha2 values and store them in a "cached matrix". 

```{r}
# Alpha array
alphaValues <- round(seq(-5, 5, by=0.1), digits=1)
cached.lcp <- array(NA, c(M, M, length(alphaValues)))

for(a in 1:length(alphaValues)){
  cached.lcp[1:M,1:M,a] <- leastcostpath(alphaValues[a]) # compute distances between sites for each defined alpha2
  print(a)}
```

### 2.3.2 Run model

```{r}
# Data and constants
my.constants <- list(nsites = M, 
                     nsurveys = J, 
                     nseasons = T)

my.data <- list(y = apply(Data[["y"]], MARGIN = c(1,3), FUN = sum),
                cached = cached.lcp,
                one = 1)

zst <- array(1, dim = c(M, T))

initial.values <- function(){ list(z = zst, 
                                   psi1 = runif(1,0,1), 
                                   sigma = rgamma(1, shape = 1, rate = 1),
                                   gamma0 = runif(1,0,1),
                                   phi = runif(1,0,1), 
                                   p = runif(1,0,1),
                                   alpha2 = runif(1,-5,5))}

# Initialization and compilation
occupancy <- nimble::nimbleModel(code = lcp.model.cached, 
                                 data = my.data,
                                 constants = my.constants,
                                 inits = initial.values(),
                                 calculate = FALSE) # first tip

Coccupancy <- nimble::compileNimble(occupancy)


occupancyConf <- nimble::configureMCMC(occupancy,
                                       enableWAIC = TRUE)

occupancyMCMC <- nimble::buildMCMC(occupancyConf, useConjugacy = FALSE) # second tip


CoccupancyMCMC <- nimble::compileNimble(occupancyMCMC, 
                                        project = occupancy)


# Run Nimble
samples <- nimble::runMCMC(mcmc = CoccupancyMCMC, 
                           niter = 10000, 
                           nburnin = 5000,
                           nchains = 2,
                           WAIC = TRUE)

# Plots
mcmc.output.lcp <- samples
mcmc.output.lcp$WAIC
MCMCvis::MCMCtrace(object = mcmc.output.lcp$samples,
                   pdf = FALSE, # no export to PDF
                   ind = TRUE, 
                   params = c("gamma0", "sigma", "psi1", "phi", "p", "alpha2"),
                   gvals = c(gamma0, sigma, psi1, phi, p, alpha2))
```

## 2.4  Commute distance from circuit theory

### 2.4.1 Compute cached matrix 

To avoid to compute commute distance at each MCMC iteration, we precalculate distance between sites for a set of alpha2 values and store them in a "cached matrix"
```{r}
# Alpha array
alphaValues <- round(seq(-5, 5, by=0.1), digits=1)
cached.circuit <- array(NA, c(M, M, length(alphaValues)))

for(a in 1:length(alphaValues)){
  cached.circuit[1:M,1:M,a] <- CircuitDistance(alphaValues[a]) # compute distances between sites for each defined alpha2
  print(a)}
```

### 2.4.2 Run model
```{r}
# Data and constants --------------------------
my.constants <- list(nsites = M, 
                     nsurveys = J, 
                     nseasons = T)

my.data <- list(y = apply(Data[["y"]], MARGIN = c(1,3), FUN = sum),
                cached = cached.circuit,
                one = 1)

zst <- array(1, dim = c(M, T))

initial.values <- function(){ list(z = zst, 
                                   psi1 = runif(1,0,1), 
                                   sigma = rgamma(1, shape = 1, rate = 1),
                                   gamma0 = runif(1,0,1),
                                   phi = runif(1,0,1), 
                                   p = runif(1,0,1),
                                   alpha2 = runif(1,-5,5))}

# Initialization and compilation
occupancy <- nimble::nimbleModel(code = circuit.model.cached, 
                                 data = my.data,
                                 constants = my.constants,
                                 inits = initial.values(),
                                 calculate = FALSE) # first tip

Coccupancy <- nimble::compileNimble(occupancy)

occupancyConf <- nimble::configureMCMC(occupancy,
                                       enableWAIC = TRUE)

occupancyMCMC <- nimble::buildMCMC(occupancyConf, useConjugacy = FALSE) # second tip


CoccupancyMCMC <- nimble::compileNimble(occupancyMCMC, 
                                        project = occupancy)

# Run Nimble
samples <- nimble::runMCMC(mcmc = CoccupancyMCMC, 
                           niter = 10000, 
                           nburnin = 5000,
                           nchains = 2,
                           WAIC = TRUE)

# Plots
mcmc.output.circuit <- samples
mcmc.output.circuit$WAIC
MCMCvis::MCMCtrace(object = mcmc.output.circuit$samples,
                   pdf = FALSE, # no export to PDF
                   ind = TRUE, 
                   params = c("gamma0", "sigma", "psi1", "phi", "p", "alpha2"),
                   gvals = c(gamma0, sigma, psi1, phi, p, alpha2))
```

# References

Chandler, R. B., E. Muths, B. H. Sigafus, C. R. Schwalbe, C. J. Jarchow, and B. R. Hossack. 2015. Spatial occupancy models for predicting metapopulation dynamics and viability following reintroduction. Journal of Applied Ecology 52:1325–1333.

Howell, P. E., E. Muths, B. R. Hossack, B. H. Sigafus, and R. B. Chandler. 2018. Increasing connectivity between metapopulation ecology and landscape ecology. Ecology 99:1119–1128.

MacKenzie, D. I., J. D. Nichols, J. A. Royle, K. H. Pollock, L. Bailey, and J. E. Hines. 2017. Occupancy Estimation and Modeling: Inferring Patterns and Dynamics of Species Occurrence. Second edition. Elsevier.

de Valpine, P., D. Turek, C. J. Paciorek, C. Anderson-Bergman, D. T. Lang, and R. Bodik. 2017. Programming with models: writing statistical algorithms for general model structures with NIMBLE. Journal of Computational and Graphical Statistics 26:403–413.